### 2019-02-26 2-Chapters/4-Chapter/IEEE_WCNC__2019__Paper__BMBM.git/IEEE_WCNC__2019__Paper__BMBM.tex
Done for now, I don't like the content of Chapter 3 at all for now, but at least there is something.
Proofreading of Chapter 4. No new science since the papers... but it's cleanly presented and unified.

% https://perso.crans.org/besson/articles/BMBM__IEEE_WCNC_2019.pdf

@@

% \TODOL{This Section is based on the publication ``Upper-Confidence Bound for Channel Selection in LPWA Networks with Retransmissions'', see \texttt{https://perso.crans.org/besson/articles/BMBM\_\_IEEE\_WCNC\_2019.pdf}}

\TODOL{TODO:

- Be sure to not repeat any motivations / maths / explanations given before
}

@@

	\item
	Secondly, several heuristics are proposed to cope with retransmissions.

@@

The rest of the Section is organized as follows.
First the system model is introduced in Section~\ref{sub:43:model},
% A formal description of the MAB learning algorithms is given in Section~\ref{sub:43:motivations},
and our motivations are exposed in Section~\ref{sub:43:MABalgo}.
The proposed \UCB-based heuristics are presented in Section~\ref{sub:43:heuristics}, while the corresponding numerical results are shown in Section~\ref{sub:43:numExp}.
Finally, some conclusions are drawn in Section~\ref{sub:43:conclusion}.

@@

Like before in this chapter, we consider an LPWA network composed of a gateway and a large number of end-devices that regularly send short data packets, where $K$ channels ($K>1$) are available for the transmission of their packets.
%
We assume that this network is constituted by two types of devices:
on the one hand, we have \emph{static} devices that operate in one channel\footnote{~Note that, for unlicensed bands, this definition also encompasses any device following a different standard or trying to establish communication with gateways of other networks.} in order to communicate with the gateway.
%
On the other hand, there are  IoT devices, that possess the additional advantage of being able to select any of the $K$ available channels to perform their transmissions.

Regardless the type of devices, each of them follows a slotted ALOHA protocol \cite{Roberts75}, and has a probability $p>0$ to transmit a packet in a time slot.
We make the hypothesis that the transmission is successful if the channel is available, otherwise upon radio collision, these devices will attempt to retransmit their packet up-to $M$ times, with $M \in\mathbb{N}$.
Note that, every retransmission is carried out after a random back-off time, uniformly distributed in $\llbracket 0, m-1 \rrbracket$, where $m \in\mathbb{N}, m>0$ is the length of the back-off interval.

@@

The aforementioned contention process can be described by a Markov chain model \cite{Norris98} similar to the one presented in \cite{Yang12}, as it is depicted in Fig.~\ref{fig:43:Markov_model}.
A device containing a packet for transmission goes from an idle state to a transmission state, while considering retransmissions due to different collision probabilities, \ie, $\{p_{c}, p_{c1}, \dots, p_{M-2} \}$, at each $M$ back-off stage.
At each time slot, a transition from an idle state to a transmission state (denoted as \texttt{Trans.}) occurs if a packet transmission is required, while waiting states (denoted as \texttt{Wait}), correspond to a $m$ back-off interval.

A device aims to select a channel with the highest probability of a successful transmission, for which it resorts to a reinforcement learning approach. It is formulated as MAB problem, where each channel (also called arms) is viewed as a gambling machine (bandit), and each bandit has a \emph{reward}. Then, at every trial, a device chooses a channel that maximizes the sum of the collected rewards. These \emph{rewards} are the \emph{acknowledgment} (\emph{Ack}) signals received after transmitting packets to the gateway. In this way, a transmission is considered successful when an acknowledgment is received, and a learning approach is employed to select the best channel.

We address the problem of channel selection taking into account the described Markov model for the retransmissions of end-devices.
% It motivates our present work for which we consider the retransmissions in the analysis of MAB algorithms.
It motivates our present work for which we consider the number of retransmissions, carried out by each device to address several MAB algorithms.

@@

We assume to be in the steady state \cite{Norris98}, in our Markov chain model depicted in Figure~\ref{fig:43:Markov_model}, and thus the probabilities no longer depend on the slot number $t$ (\ie, $\forall t, x_t=x$).
Therefore, the probability that this device has a collision at the first transmission is $p_c$, and has the following expression
\begin{equation}\label{eq:43:1}
	p_c = 1-\left(1-x_t\right)^{N-1} \iff x_t = 1-\left(1-p_c\right)^{\frac{1}{N-1}}.
\end{equation}

@@

\TODOL{Remove this part or compress}

@@

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 0, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission}){
				Compute for each channel $U^r_k(t) = \widehat{\mu^r_k}(t) + B^r_k(t)$ following Eqs.~\eqref{eq:43:Nkt},~\eqref{eq:43:mukt}, and~\eqref{eq:43:Bkt}\;
				Transmit in channel $C^r(t) = \arg\max_{1\leq k \leq K} U^r_k(t)$\;
				Reward $r^r_{C^r(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Heuristic: \UCB{} for retransmission.}    % A naive
		\label{algo:43:TwoUCB}
		% \end{small} % XXX remove if needed
		\end{algorithm}
\end{figure}
% \end{small} % XXX remove if needed

@@

Although, this approach increases the complexity and store requirements (of order $\mathcal{O}(K^2)$).
For our LPWA networks of interest, such as LoRaWAN, the cost of its implementation is still affordable, since a small number of channels is used.
For instance, for $K=4$ channels,
the memory to store $K+1=5$ algorithms is of the order of the requirements to storing one.

@@

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
	\begin{figure}[h!]
		\centering
		\begin{algorithm}[H]
			% \begin{small} % XXX remove if needed
			% \KwData{$\forall k,j\in[\![1;K]\!]$, $N_k^j(t)$, $\widehat{\mu_k}^j(t)$, $B_k^j(t)$ and $U_k^j(t)$}
			\For(\tcp*[f]{At every time step})
			{$t = 0, \dots, T$}{
				\uIf{First packet transmission}{
					Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
				}
				\Else(\tcp*[f]{Packet retransmission}){ % Retr. after trying channel $j$
					j $\leftarrow$ last channel selected by first-stage \UCB\;
					Compute for each channel $U_k^j(t) = \widehat{\mu_k}^j(t) + B_k^j(t)$ following Eqs.~\eqref{eq:43:Nkt},~\eqref{eq:43:mukt}, and~\eqref{eq:43:Bkt}\;
					Transmit in channel $C^j(t) = \arg\max_{1\leq k \leq K} U^j_k(t)$\;
					Reward $r^j_{C^j(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
				}
			}
			\caption{Heuristic: $K$ different {\UCB}s for retransmission.}
			\label{algo:43:UCBthenKp1}
			% \end{small} % XXX remove if needed
		\end{algorithm}
	\end{figure}
% \end{small} % XXX remove if needed

@@

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
	\begin{figure}[h!]
		\centering
		\begin{algorithm}[H]
			% \begin{small} % XXX remove if needed
			\For(\tcp*[f]{At every time step})
			{$t = 0, \dots, T$}{
				\uIf{First packet transmission}{
			 		Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
				}
				\uElseIf(\tcp*[f]{Random retransmission}){$t \leq \Delta$}{
					Transmit randomly in a channel $C(t) \sim \mathcal{U}(1,\ldots,K)$.
				}
				\Else(\tcp*[f]{Delayed \UCB{}}){
					Compute for each channel $ U^d_k(t) = \widehat{\mu^d_k}(t) + B^d_k(t)$ following Eqs.~\eqref{eq:43:Nkt},~\eqref{eq:43:mukt}, and~\eqref{eq:43:Bkt}\;
					Transmit in channel $C^d(t) = \arg\max_{1\leq k \leq K} U^d_k(t)$\;
					Reward $r^d_{C^d(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
				}
			}
			\caption{Heuristic: Delayed \UCB{} for retransmission.}
			\label{algo:43:UCBwithDelay}
			% \end{small} % XXX remove if needed
		\end{algorithm}
	\end{figure}
% \end{small} % XXX remove if needed

% ----------------------------------------------------------------------
\subsection{Simulations to compare our heuristics}
\label{sub:43:numExp}
% ----------------------------------------------------------------------

We simulate our network considering $N$ devices following the contention Markov process described in Section \ref{sub:43:model}, and a LoRa standard with $K=4$ channels.
Each device is set to transmit with a fixed probability $p=10^{-3}$, \ie, a packet about every $20$ minutes for time slots of $1\;\mathrm{s}$.
%
For the evaluation of the proposed heuristics, a total number of $T=10^{6}$ time slots is considered, and the results are averaged over $10^{3}$ independent random simulations.
%In this way, it allows us to evaluate the learning rate, as well as the convergence of each learning approach.

@@

We also consider in our analysis the case where $M=5$, and $m=10$ using ALOHA protocol, a statistic distribution of the devices about $40\%, 30\%, 20\%, 10\%$ for the four channels, and $N=2000$ IoT devices.
The corresponding results are depicted in Fig.~\ref{fig:43:mainExperiment2}.
In this case the successful transmission rate is degraded compared with achieved results in Fig.~\ref{fig:43:mainExperiment1}, this can be explained with the fact that we are considering in our network more devices that increase the collision probability.
It is important to highlight,that the ``Random" retransmission heuristic shows a poor performance in comparison to the other heuristics, and it can be attributed to the fact that the number of retransmission is increased, and consequently a
learning approach is able to take advantage of it.
Furthermore, the ``Delayed \UCB'' and  the``\UCB'' heuristics show quite similar results, after converging.
%
%Random, \UCB{}, $K$ \UCB{}, Delayed \UCB{} obtain similar results. The less efficient one is the $K$ \UCB{}, and this makes sense as having more learning algorithm needs more for each of them to learn. The most efficient one is the simple \UCB{} procedure, and this was quite surprising.

>>>>

% - ``Upper-Confidence Bound for Channel Selection in LPWA Networks with Retransmissions'', see  https://hal.inria.fr/hal-02049824

@@

% \TODOL{This Section is based on the publication ``Upper-Confidence Bound for Channel Selection in LPWA Networks with Retransmissions'', see \texttt{https://hal.inria.fr/hal-02049824}}

% \TODOL{TODO:

% - Include the modifications suggested by the reviewers of our paper!
% - Be sure to not repeat any motivations / maths / explanations given before
% }

% https://edas.info/showPaper.php?m=1570524364
% https://hal.inria.fr/hal-02049824

@@

	\item
	Secondly, different heuristics are proposed to cope with retransmissions.

@@

The rest of the Section is organized as follows.
First the system model is introduced in Section~\ref{sub:43:model},
% A formal description of the MAB learning algorithms is given in Section~\ref{sub:43:motivations},
and our motivations are exposed in Section~\ref{sub:43:MABalgo}.
The proposed \UCB-based heuristics are presented in Section~\ref{sub:43:heuristics}, while the corresponding numerical results are shown in Section~\ref{sub:43:numExp}.
Finally, some conclusions are drawn later in Section~\ref{sub:43:conclusion}.

@@

Like before in this chapter, we consider an LPWA network composed of a gateway and a large number of end-devices that regularly send short data packets, where $K$ channels ($K>1$) are available for the transmission of their packets.
%
We assume that this network is constituted by two types of devices:
on one hand, we have \emph{static} devices that operate in one channel\footnote{~Note that, for unlicensed bands, this definition also encompasses any device following a different standard or trying to establish communication with gateways of other networks.} in order to communicate with the gateway.
%
On the other hand, there are  IoT devices, that possess the additional advantage of being able to select any of the $K$ available channels to perform their transmissions.

Like in the previous model presented in Section~\ref{sec:4:firstModel},
regardless the type of devices, each of them follows a slotted ALOHA protocol \cite{Roberts75}, and has a probability $p>0$ to transmit a packet in a time slot.
We make the hypothesis that the transmission is successful if the channel is available, otherwise upon radio collision, these devices will attempt to retransmit their packet up-to $M$ times, with $M \in\mathbb{N}$.
Note that, every retransmission is carried out after a random back-off time, uniformly distributed in $\llbracket 0, m-1 \rrbracket$, where $m \in\mathbb{N}, m>0$ is the length of the back-off interval.

@@

The aforementioned contention process can be described by a Markov chain model \cite{Norris98} similar to the one presented in \cite{Yang12}, as it is depicted in Fig.~\ref{fig:43:Markov_model}.
When a device have a packet to transmit, it goes from an idle state to a transmission state, while considering retransmissions due to different collision probabilities, \ie, $\{p_{c}, p_{c1}, \dots, p_{M-2} \}$, at each $M$ back-off stage.
At each time slot, a transition from an idle state to a transmission state (denoted as \texttt{Trans.}) occurs if a packet transmission is required, while waiting states (denoted as \texttt{Wait}), correspond to a $m$ back-off interval.

A device aims to select a channel with the highest probability of a successful transmission, for which it resorts to a reinforcement learning approach, formulated as a MAB problem.
% , where each channel (also called arms) is viewed as a gambling machine (bandit), and each bandit has a \emph{reward}. Then, at every trial, a device chooses a channel that maximizes the sum of the collected rewards. These \emph{rewards} are the \emph{acknowledgment} (\emph{Ack}) signals received after transmitting packets to the gateway. In this way, a transmission is considered successful when an acknowledgment is received, and a learning approach is employed to select the best channel.
%
We address the problem of channel selection taking into account the described Markov model for the retransmissions of end-devices.
% It motivates our present work for which we consider the retransmissions in the analysis of MAB algorithms.
It motivates our present work for which we consider the number of retransmissions, carried out by each device to address several MAB algorithms.

@@

We assume to be in the steady state \cite{Norris98}, in our Markov chain model depicted in Figure~\ref{fig:43:Markov_model}, and thus the probabilities no longer depend on the slot number $t$ (\ie, $\forall t, x_t=x$).
Therefore, the probability that this device has a collision at the first transmission is $p_c$, and has the following expression
\begin{equation}\label{eq:43:1}
	p_c = 1-\left(1-x\right)^{N-1} \iff x = 1-\left(1-p_c\right)^{\frac{1}{N-1}}.
\end{equation}

@@

\TODOL{Remove this part or compress: I will remove as soon as I have clear equations in Chapter~\ref{chapter:2} as references for the \UCB{} algorithm.}

@@

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 0, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^r$}){
				Compute for each channel $U^r_k(t) = \widehat{\mu^r_k}(t) + B^r_k(t)$ following Eqs.~\eqref{eq:43:Nkt},~\eqref{eq:43:mukt}, and~\eqref{eq:43:Bkt}\;
				Transmit in channel $C^r(t) = \arg\max_{1\leq k \leq K} U^r_k(t)$\;
				Reward $r^r_{C^r(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Heuristic: \UCB{} for retransmission.}    % A naive
		\label{algo:43:TwoUCB}
		% \end{small} % XXX remove if needed
		\end{algorithm}
\end{figure}
% \end{small} % XXX remove if needed

@@

Although, this approach increases the complexity and storage requirements (of order $\mathcal{O}(K^2)$).
For our LPWA networks of interest, such as LoRaWAN, the cost of its implementation is still affordable, since a small number of channels is used.
For instance, for $K=4$ channels,
the memory to store $K+1=5$ algorithms is of the order of the requirements to store one.

@@

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
	\begin{figure}[h!]
		\centering
		\begin{algorithm}[H]
			% \begin{small} % XXX remove if needed
			% \KwData{$\forall k,j\in[\![1;K]\!]$, $N_k^j(t)$, $\widehat{\mu_k}^j(t)$, $B_k^j(t)$ and $U_k^j(t)$}
			\For(\tcp*[f]{At every time step})
			{$t = 0, \dots, T$}{
				\uIf{First packet transmission}{
					Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
				}
				\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^j$}){ % Retr. after trying channel $j$
					j $\leftarrow$ last channel selected by first-stage \UCB\;
					Compute for each channel $U_k^j(t) = \widehat{\mu_k}^j(t) + B_k^j(t)$ following Eqs.~\eqref{eq:43:Nkt},~\eqref{eq:43:mukt}, and~\eqref{eq:43:Bkt}\;
					Transmit in channel $C^j(t) = \arg\max_{1\leq k \leq K} U^j_k(t)$\;
					Reward $r^j_{C^j(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
				}
			}
			\caption{Heuristic: $K$ different {\UCB}s for retransmission.}
			\label{algo:43:UCBthenKp1}
			% \end{small} % XXX remove if needed
		\end{algorithm}
	\end{figure}
% \end{small} % XXX remove if needed

@@

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
	\begin{figure}[h!]
		\centering
		\begin{algorithm}[H]
			% \begin{small} % XXX remove if needed
			\For(\tcp*[f]{At every time step})
			{$t = 0, \dots, T$}{
				\uIf{First packet transmission}{
			 		Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
				}
				\uElseIf(\tcp*[f]{Random retransmission with $\mathrm{UCB}^d$}){$t \leq \Delta$}{
					Transmit randomly in a channel $C(t) \sim \mathcal{U}(1,\ldots,K)$.
				}
				\Else(\tcp*[f]{Delayed \UCB{}}){
					Compute for each channel $ U^d_k(t) = \widehat{\mu^d_k}(t) + B^d_k(t)$ following Eqs.~\eqref{eq:43:Nkt},~\eqref{eq:43:mukt}, and~\eqref{eq:43:Bkt}\;
					Transmit in channel $C^d(t) = \arg\max_{1\leq k \leq K} U^d_k(t)$\;
					Reward $r^d_{C^d(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
				}
			}
			\caption{Heuristic: Delayed \UCB{} for retransmission.}
			\label{algo:43:UCBwithDelay}
			% \end{small} % XXX remove if needed
		\end{algorithm}
	\end{figure}
% \end{small} % XXX remove if needed

% ----------------------------------------------------------------------
\subsection{Numerical results}
\label{sub:43:numExp}
% ----------------------------------------------------------------------

We simulate our network considering $N$ devices following the contention Markov process described in Section \ref{sub:43:model}, and a LoRa standard with $K=4$ channels.
Each device is set to transmit with a fixed probability $p=10^{-3}$, \ie, a packet about every $20$ minutes for time slots of $1\;\mathrm{s}$.
%
For the evaluation of the proposed heuristics, a total number of $T=20 \times 10^4$ time slots is considered, and the results are averaged over $10^{3}$ independent random simulations.
%In this way, it allows us to evaluate the learning rate, as well as the convergence of each learning approach.

@@

We also consider in our analysis the case where $M=5$, and $m=10$ using ALOHA protocol, a statistic distribution of the devices about $40\%, 30\%, 20\%, 10\%$ for the four channels, and $N=2000$ IoT devices.
The corresponding results are depicted in Fig.~\ref{fig:43:mainExperiment2}.
In this case the successful transmission rate is degraded compared with achieved results in Fig.~\ref{fig:43:mainExperiment1}, this can be explained with the fact that we are considering in our network more devices that increase the collision probability.
It is important to highlight,that the ``Random" retransmission heuristic shows a poor performance in comparison to the other heuristics, and it can be attributed to the fact that the number of retransmission is increased, and consequently a
learning approach is able to take advantage of it.
Furthermore, the ``\UCB'', ``$K$ \UCB'' and ``Delayed \UCB'' heuristics behave similarly than ``Only \UCB'', after a similar convergence time.

### 2019-02-26 2-Chapters/4-Chapter/chapter4.tex
Proofreading of Chapter 4. No new science since the papers... but it's cleanly presented and unified.

In this Chapter~\ref{chapter:4}, we present two models of Internet of Things networks, composed of many independent devices that are able to run MAB algorithm to learn and improve their access to the network.
The two models are presented mathematically, and numerical simulations justify the interest of the proposed algorithms.
We also present a proof-of-concept real-world validation of the advantage of using Multi-Armed Bandit algorithms for the first model, by showing and explaining a demonstration implemented on USRP \cite{USRPDocumentation} cards.

More precisely, on the one hand, we start by presenting in Section~\ref{sec:4:firstModel} a model where independent devices use an IoT network composed of different channels, used by static devices.
Devices do not use any sensing, but use acknowledgements sent by the Base Station to know if their communication messages were successfully received.
We consider the case when many independent ``dynamic'' devices all have a small probability of communicating at every discrete time. If they use simple MAB algorithms in a decentralized and independent fashion (what we call ``selfish'' learning),
we show in simulations that in the case of non-uniform repartition of static devices in the channels, the dynamic devices can improve a lot the network efficiency.
%
On a simple example, we present in Section~\ref{sec:4:gnuradio} a demonstration that implements and validates the proposed approach, using real radio hardware.

On the other hand, we present in Section~\ref{sec:4:retransmissions} a second model extending the first one, when devices can retransmit their messages for a fixed number of maximum retransmission (like in the ALOHA protocol).
In this second model, we compare different heuristic, all based on the UCB algorithm, and numerical simulation show that all the non-naive heuristics succeed in significantly improving the network efficiency.

@@

Efficient Medium Access (MAC) policies allow devices to avoid interfering traffic and can significantly reduce the spectrum contention problem in unlicensed bands.
As end-devices battery life is a key constraint of IoT networks,
this leads to IoT protocols using as low signaling overhead as possible and simple ALOHA-based mechanisms.
%
In this Section, we analyze the performance of Multi-Armed Bandits (MAB) algorithms, used in combination with a time-frequency slotted ALOHA-based protocol.
We consider the Upper-Confidence Bound (\UCB) \cite{Auer02}, and the Thompson-Sampling (TS) algorithms \cite{Thompson33,AgrawalGoyal11,
Kaufmann12Thompson}.

MAB learning has already been proposed in Cognitive Radio (CR) \cite{Haykin05}, and in particular, for sensing-based Dynamic Spectrum Access (DSA) in licensed bands \cite{Jouini10}.
For example, \cite{Jouini10} presents the OSA setting, and \cite{Maghsudi16} shows how MAB learning can be applied for small cell management in licensed 5G networks.
Recently, TS and \UCB{} algorithms have been used for improving the spectrum access in (unlicensed) WiFi networks \cite{Toldov16}.
Many recent works show that MAB algorithms work well for real-world radio signal.
However, even with only one dynamic user using the learning algorithm, the background traffic or the traffic of the other devices is never really stationary or \iid{}.

We present in Section~\ref{sec:4:firstModel} how the MAB algorithms can be used in a unlicensed and frequency- and time-slotted IoT network \cite{Bonnefoi17}.
Several devices are using bandit algorithms, and the assumptions made by the stochastic bandit algorithms are not satisfied: as several agents learn simultaneously, their behavior is neither stationary nor \iid.
As far as we know, we provide the first study to confirm robustness of the use of stochastic bandit algorithms for decision making in IoT networks with a large number of intelligent devices in the network, which makes the environment not stationary at all, violating the hypothesis required for mathematical proofs of bandit algorithms convergence and efficiency.
We then validate the model with a hardware implementation, detailed in Section~\ref{sec:4:gnuradio}.
%
We conclude this Chapter by presenting an extension of this model to take into account another aspect of the ALOHA protocol, that is the possibility for dynamic devices to retransmit their packet if the \emph{Ack} was not received.

>>>>

In this Chapter~\ref{chapter:4}, we present two models of Internet of Things networks, composed of many independent devices that are able to run MAB algorithm to learn and improve their access to the network.
The two models are presented mathematically, and numerical simulations justify the interest of the proposed algorithms.
We also present a proof-of-concept real-world validation of the advantage of using Multi-Armed Bandit algorithms for the first model, by showing and explaining a demonstration implemented on USRP cards \cite{USRPDocumentation}.

More precisely, on one hand, we start by motivating this idea of independent ``selfish'' learning for IoT devices in Section~\ref{sec:4:motivations}.
We then present in Section~\ref{sec:4:firstModel} a model where independent devices use an IoT network composed of different channels, used by static devices.
Devices do not use any sensing, but use acknowledgements sent by the Base Station to know if their communication messages were successfully received.
We consider the case when many independent ``dynamic'' devices all have a small probability of communicating at every discrete time. If they use simple MAB algorithms in a decentralized and independent fashion (what we call ``selfish'' learning),
we show in simulations that in the case of non-uniform repartition of static devices in the channels, the dynamic devices can improve a lot the network efficiency.
%
On a simple example, we present in Section~\ref{sec:4:gnuradio} a demonstration that implements and validates the proposed approach, using real radio hardware.

On the other hand, we present in Section~\ref{sec:4:retransmissions} a second model extending the first one, when devices can retransmit their messages for a fixed number of maximum retransmission (like in the ALOHA protocol).
In this second model, we compare five different heuristics, all based on the UCB algorithm, and numerical simulation show that all the non-naive heuristics succeed in significantly improving the network efficiency.
Interestingly, the simpler heuristic is found to be the most efficient one, as its simplicity implies a faster learning and convergence time.

@@

Efficient Medium Access (MAC) policies allow devices to avoid interfering traffic and can significantly reduce the spectrum contention problem in unlicensed bands.
As end-devices battery life is a key constraint of IoT networks,
this leads to IoT protocols using as low signaling overhead as possible and simple ALOHA-based mechanisms.
%
In this Chapter, we analyze the performance of Multi-Armed Bandits (MAB) algorithms, used in combination with a time-frequency slotted ALOHA-based protocol.
We consider the Upper-Confidence Bound (\UCB) \cite{Auer02}, and the Thompson-Sampling (TS) algorithms \cite{Thompson33,AgrawalGoyal11,
Kaufmann12Thompson}, for the first model. For the demonstration and the second model, we preferred to focus on heuristics based on the simplest algorithm (\ie, \UCB), to give a clear presentation of the different ideas explored for to solve the problem of learning to retransmit efficiently.

As detailed in Chapter~\ref{chapter:1},
MAB learning has already been proposed in Cognitive Radio (CR) \cite{Haykin05}, and in particular, for sensing-based Dynamic Spectrum Access (DSA) in licensed bands \cite{Jouini10}.
% For example, \cite{Jouini10} presents the OSA setting, and \cite{Maghsudi16} shows how MAB learning can be applied for small cell management in licensed 5G networks.
Recently, TS and \UCB{} algorithms have been used for improving the spectrum access in (unlicensed) WiFi networks, for instance by \cite{Toldov16}.
Many recent works show that MAB algorithms work well for real-world radio signal.
However, even with only one dynamic user using the learning algorithm, the background traffic or the traffic of the other devices is never really stationary or \iid{}.

We present in Section~\ref{sec:4:firstModel} how the MAB algorithms can be used in a unlicensed and frequency- and time-slotted IoT network \cite{Bonnefoi17}.
Several devices are using bandit algorithms, and the assumptions made by the stochastic bandit algorithms are not satisfied: as several agents learn simultaneously, their behavior is not stationary.
As far as we know, we provide the first study to confirm robustness of the use of stochastic bandit algorithms for decision making in IoT networks with a large number of intelligent devices in the network, which makes the environment not stationary at all, violating the hypothesis required for mathematical proofs of bandit algorithms convergence and efficiency.
We then validate the model with a hardware implementation, detailed in Section~\ref{sec:4:gnuradio}.
%
We conclude this Chapter by presenting an extension of this model to take into account another aspect of the ALOHA protocol, that is the possibility for dynamic devices to retransmit their packet if the \emph{Ack} was not received.
